Servis: A dependently typed DSL for web APIs

### 2016-08-25 paper/paper.tex
Write more and more

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{minted}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\author{Arian van Putten \\ a.vanputten@uu.nl}
\title{Servis: A dependently typed DSL for web APIs}
\begin{document}
\maketitle
\begin{abstract}
\emph{TODO}
%In this paper, we describe a DSL for web APIs named Servis. Servis is heavily inspired by Servant, a haskell type-level DSL for web APIs,  but tries to solve identified shortcomings by using a dependently typed language called Idris. Servis uses the universe pattern to deduce types from an API description such that we can verify at compile time that the implementation of the API matches the description.

%Furthermore, we take a look at if dependent types can bring more to the table in terms of expressiveness of API descriptions than the original Servant DSL\@. Specifically, we introduce Sigma-types to describe interdependencies within an API description.

% TODO: rewrite abstract \emph{after} I finish the paper :-)

@@

\section{Introduction}
\emph{TODO expand and reword}

Servant is an extensible type-level DSL for Haskell, that tries to make web APIs \emph{first-class}~\cite{servant}. A web API should be a thing that can be named, passed to, or returned by functions. Just like any other term in a language.

It does this by describing the DSL at the type-level. And then uses \emph{type families} (type-level functions) to be able to perform computations on the API specifications.  Using these type families, we are able to write different \emph{interpretations} of an API\@.  Example of interpretations are: generating documentation from a specification, generating types of a server implementation, or even generating a client that can interfact with the API\@.

In this paper, I will describe a port of Servant in Idris, and explore what kind of benefits this brings us.

\subsection{An introduction to Servant}
I will now give a very short overview of the internals of Servant. We will start by describing the type-level DSL of servant.

@@

data Get (ctypes :: [*]) (t :: *)
data Post (ctypes :: [*]) (t :: *)
data JSON
\end{minted}
\caption{The Servant DSL}
\label{lab:servant}
\end{listing}

\emph{TODO give a short overview of servant}

Servant has a few limitations which we will describe in Section~\ref{sec:limitations}.

\subsection{Why idris}
Idris is a language that in syntax is rather similar to Haskell, but has a more powerful type system. Idris is a dependently-typed language. This means that types are fist-class citizens. We can pass types to functions, let functions create new types, and let types depend on other value-level terms.

Though haskell has type-level functions (called type-families), it does not allow these functions to dpend on term-level values. There is a distinct boundary between type-level and term-level, whilst the two are merged in Idris. This makes doing computations with types a lot more natural (no special constructs are needed, just write functions), but also lets us do some more powerful stuff like letting types actually depend on runtime user input.

@@

TODO reword
\begin{itemize}
  \item I point out several problems an limitations of the original Servant DSL
  \item I show a port of servant to Idris, a dependently-typed language
  \item I show how new techniques from Idris can help mitigate some of the problems that Servant has.
\end{itemize}

\section{Limitations of Servant}\label{sec:issues}
\subsection{Term and type-level are not unified in Haskell}
As explained in the introduction, the Servant DSL lives in the type-level. In haskell, types and terms are separated. One can lift terms into the type-level using GHC extensions like \texttt{XDataKinds}, but that only gets us so far.  On the term level, Haskell has many great libraries and data structures that can be used. For example, all kinds of string manipulations. It would be useful to be able to use these functions to manipulate APIs, making them truly first-class. Now they are only first-class in the type level.

For example, a useful function would be \mintinline{haskell}{path "/users/friends/etc" = "users" :> "friends" :> "etc"} to reduce noise in the DSL\@. But this is currently impossible to write because Strings in the type level are of kind \mintinline{haskell}{Symbol} and not \mintinline{haskell}{String} so all our favorite string manipulation functions like \mintinline{haskell}{splitOn "/"} will not work. This would mean that we would have to re-implement all kinds of utility functions on the type-level.

\subsection{A kind of APIs}
As mentioned in Section~\ref{sec:servant_intro}, a delibirate choice has been made to let the Servant DSL live in the open kind \mintinline{haskell}{*}. Though this makes the DSL extensible, it also makes the DSL fragile and prone to error.

@@

The problem is hard to spot, but the user typed wrote \mintinline{haskell}{->} instead of \mintinline{haskell}{:>}.  This code compiles fine though. Because the Servant DSL lives in kind \mintinline{haskell}{*}, \mintinline{haskell}{:>} is   of kind \mintinline{haskell}{* -> * -> *}. Which is the same kind as the kind of \mintinline{haskell}{->}.

@@

Also, as described in Section~\ref{sec:servant_intro}, the Servant DSL is split up in several semantic portions.  For example, \mintinline{haskell}{:>} should take an \mintinline{haskell}{item} on the left-hand side, which describes either a constraint like \mintinline{haskell}{ReqBody} or a path-piece, whilst the right-hand side should be an \mintinline{haskell}{api}. These semantic portions are implict and not checked though. So we can create totally non-sensical API types like \mintinline{haskell}{type Nonsense = (Int :>Int) :<|> ("hey ":> Int)} that have no interpretation at all.

@@

Another implict rule is, that a chain of \mintinline{haskell}{:>} should always end in a \mintinline{haskell}{Get} or a \mintinline{haskell}{Post}. So this should be invalid: \mintinline{haskell}{type API = "a" :> ReqBody [JSON] User}.  This can also not be checked if we do not add more fine-grained kinds to the DSL.

\subsection{API Interdependencies}
\label{sec:inter}
Sometimes, one one like to describe even more type-safety than currently possible in Servant. Maybe we want to let the output-type of a request depend on a query parameter. A use case could be the following: We have a route \texttt{"/users/list?limit=100}", and we want to make sure that a handler never returns more than 100 users.  Thus the return type of the handler should be a \mintinline{idris}{BoundedList 100 User}. Or more generally, the return type should be \mintinline{idris}{BoundedList limit User}, where \mintinline{idris}{limit : Nat} comes from parsing the URL.  The full type of the API will thus depend on some runtime value. This is a textbook example of dependent types, and is currently not possible to implement in Haskell. However, with idris this should be a breeze.

\subsection{Overlapping routes}
Another problem with Servant is that there is no protection for overlapping routes. If two routes could resolve to the same path, Servant makes a left-biased choice~\cite{servant}.
It would be nice if we could verify at compile time that no routes in an API specification can overlap and give the user an error if he violates the overlapping constraint.

\subsection{Explicit status codes}
An API that describes how stuff goes right is good, and API that describes how stuff goes wrong is great. But currently, Servant has no way to indicate in the API description what kind of status codes an endpoint can return. Thus this also does not show up in the generated documentation.

It would be nice to add a list of possible status codes that an endpoint can returns, and then possibly also be able to check whether a server implementation actually only returns one of the specified status codes if something goes wrong (or right).

\begin{listing}
\begin{minted}{haskell}
  type API  = "users" :> Capture "args" String :> Get [JSON] User
         :<|> "users" :> "favorites" :> GET [JSON] User
\end{minted}{haskell}
\caption{An example of overlapping routes. The capture will be chosen if args equals "favorites"}
\end{listing}

@@

 In idris, we could define this as the following interface:
 \begin{listing}
 \begin{minted}{idris}
interface Universe (u : Type) where
  el : u -> Type
 \end{minted}{idris}
   \caption{Universe interface in Idris}
 \end{listing}

@@

--  el api evaluates to  Int -> IO User
handler : el api
handler = getUser
 \end{minted}
\subsection{The DSL}
So what does this type \mintinline{idris}{API} look like? An API DSL should allow us to define routes, to which we can attach handler types.
\subsubsection{Handler}\label{sec:handler}
Before we will look at routes, lets think of what an endpoint should look like.
We should be able to handle the HTTP verbs. We will only implement \mintinline{idris}{GET} and \mintinline{idris}{POST}, because the others are very similar.
For a \mintinline{idris}{GET}, we should be able to specify a return type of the handler,  and for a \mintinline{idris}{POST}, we should be able to
specify a request body type, and a return type.

@@

However, once we start implementating an actual server in Section~\ref{sec:server}, we will discover that this definition is not ideal. What if we want to write some
function that does something different based on the \mintinline{idris}{responseType} of a \mintinline{idris}{GET}?
For example, based on value of \mintinline{idris}{responseType}, we need to decide how to encode the response body to JSON. Lets imagine a hypothetical function \mintinline{idris}{selectEncoder} that tries to implement this behaviour:

@@

userEncoder : Encoder
userEncoder = selectEncoder User
\end{minted}{idris}

Sadly enough, this will give us a compiler error. The problem is that in Idirs, we cannot pattern match on values of \mintinline{idris}{Type}. 
However, this exact problem can be solved with universes! We can define a data type \mintinline{idris}{ResponseUniverse : Type} which names
the \mintinline{idris}{User} type. We can then pattern-match on \mintinline{idris}{ResponseUniverse} instead of \mintinline{idris}{Type}!
And whenever we need the \mintinline{idris}{User} type, we can just call \mintinline{idris}{el USER : Type}.

@@

This leads us to the following definition of \mintinline{idris}{Handler}:

\begin{listing}
  \begin{minted}{idris}
data Handler : (req : Type)  -> (res : Type) -> Type where
  GET : (responseType : res) -> Handler req res
  POST : (requestType : req) -> (responseType : res) -> Handler req res

@@

\subsubsection{Path and PathPart}
Of course we do not only get request information from the request body,
we also get information about a request by parsing it's request URL. This
is where \mintinline{idris}{Path}s and \mintinline{idris}{PathPart}s come into play. 
When we talk about path parts, we mean sections of an path template that convey parsable information.
For example, the following path template has three path parts: \texttt{/users/\string{user\_id\string}?search=\string{query\string}}. A constant piece
in the path \texttt{users}, a variable capture \texttt{user\_id} and a query param \texttt{search}. This
decomposition of a path template can be written down as a datatype. In this datatype we do not only
want to describe the names of the parts of these paths, but also their types. Hence, we will parameterize
over a universe similarly to how we defined \mintinline{idris}{Handler}.
\begin{minted}{idris}

||| PathPart is a part of a path.
||| @ capture the universe of captures
||| @ query   the universe of query params
data PathPart : (capture : Type) -> (query : Type) -> Type where
  ||| A constant piece of text
  ||| @ path the path part
  Const : (path : String) -> PathPart capture query
  ||| A capture of a variable name of a specific type
  ||| @ name  the name of the variable
  ||| @ type  the type of the variable
  Capture : (name : String) -> (type : capture) -> PathPart capture query
  ||| A query param of a variable name of a specific type
  ||| @ name  the name of the variable
  ||| @ type  the type of the variable
  QueryParam : (name : String) -> (type : query) -> PathPart capture query

@@

The Path we just described can now be written as
\begin{minted}{idris}
   Const "users"
:> Capture "user_id" INT
:> QueryParam "search" STRING
:> Outputs (GET USER)
\end{minted}

The universe implementation is straightforward for paths, except for
that we do not wannt to have spurious arguments in case when we encounter
a \mintinline{idris}{Const}:
\begin{minted}{idris}
( Universe capture
, Universe query
) => Universe (PathPart capture query) where
  el (Const path) = ()
  el (Capture name type) = el type
  el (QueryParam name type) = el type

@@

\subsubsection{The API Datatype}
Finally, we define an \mintinline{idris}{API} as an non-empty list of \mintinline{idris}{Path}s.

@@

The universe implementation of an \mintinline{idris}{API} is the type of a list of handlers.
Every handler is of a different type though, so instead of using an ordinary list,
we use an heterogenous list, where each element \mintinline{idris}{x} in \mintinline{idris}{xs}
is of type \mintinline{idris}{el x}.
\begin{minted}
( Universe capture
, Universe query
, Universe req
, Universe res
) => Universe (API capture query req res) where
  el (OneOf xs) = HVect (map el xs)
\end{minted}.

\subsection{An example API}
Using this DSL, we will now describe an API for interacting with users in a database.
We will start of with defining our domain objects, and intercations we could do with them.

\begin{listing}
\begin{minted}{idris}
record User where
  constructor MkUser
  id : Int
  name : String
  email : String
  password : String

findUserByEmail : String -> IO (List User)
getUsers : (limit : Nat) -> IO (List User)
getUserById : (id : Int) -> IO User
addUser : User -> IO User
\end{minted}
\label{code:domain}
\caption{The User object and its interactions}
\end{listing}

Next, we will describe the HTTP API for interacting with these users.  This includes
the need of coming up with universes for captures, query params, request body, and response.

\begin{listing}
\begin{minted}{idris}
data QueryUniverse
  = QSTRING
  | QINT
  | QNAT
Universe QueryUniverse where
  el QSTRING = String
  el QNAT = Nat

data CaptureUniverse
  = CINT
Universe CaptureUniverse where
  el CINT = Int

@@

userAPI : API QueryUniverse CaptureUniverse ReqResUniverse ReqResUniverse
userAPI = OneOf
  [ Const "users" :> Capture "user_id" CINT :> Outputs (GET USER)
  , Const "users" :>
    Const "find" :>
    QueryParam "email" QSTRING :>
    Outputs (GET (LIST USER))
  , Const "users" :> Outputs (POST USER)
  , Const "users" :>
    QueryParam "limit" QNAT :>
    Outputs (GET (LIST USER))
  ]

-- set the handlers
handlers : el userAPI
handlers =
  [ getUserById
  , findUserByEmail
  , addUser
  , getUsers
  ]

@@

\subsection{A simple interpreter: Documentation generation}
Writing an interpreter for generating documentation is very easy. Our API description isn't some type-level definition, it's just a plain old datatype that can be passed to functions and can be manipulated.

We start by introducing an interface for docs:
\begin{minted}{idris}
interface HasDocs a where
  docs :: a -> String
\end{minted}
Now, we just define an implementation for \mintinline{idris}{API} for this interface and we're done. Which is just calling \mintinline{idris}{docs} recursively on all its children.

\begin{minted}{idris}
(HasDocs req, HasDocs res) => HasDocs (Handler req res) where
  docs (GET responseType) =
    "## GET\nResponse type: " ++ docs responseType

  docs (POST requestType responseType) =
    "## POST\nRequest type: "  ++ docs requestType ++ "\n" ++
    "Response type: " ++ docs responseType ++ "\n"

renderParams : HasDocs query => List (String, query) -> String
renderParams = renderParams' . reverse
  where
    renderParams' : List (String, query) -> String
    renderParams' [] = ""
    renderParams' ((a, b) :: []) =
      "?" ++ a ++ "=<"++ docs b ++ ">"
    renderParams' ((a, b) :: xs) =
      renderParams' xs ++ "&" ++ a ++ "=<"++ docs b ++ ">"

renderPath : 
  ( HasDocs capture
  , HasDocs query
  , HasDocs req
  , HasDocs resp) => Path capture query req resp -> String -> String
renderPath (Outputs x) params = 
    params ++ "\n" ++ docs x
renderPath ((Const x) :> y) params =
    "/" ++ x ++ renderPath y params
renderPath ((Capture name type) :> y) params =
  "/<" ++ name ++ ":" ++ docs type ++ ">" ++ renderPath y params
renderPath _ _ = ""

( HasDocs capture
, HasDocs query
, HasDocs req
, HasDocs resp
) => HasDocs (Path capture query req resp) where
  docs path =
    "# " ++ renderPath path (renderParams . getParams $ path) ++ "\n"
    where

      getParams : Path capture query req resp -> List (String, query)
      getParams (Outputs x) = []
      getParams ((Const x) :> y) = getParams y
      getParams ((Capture name type) :> y) = getParams y
      getParams ((QueryParam name type) :> y) = (name, type) :: getParams y

( HasDocs capture
, HasDocs query
, HasDocs req
, HasDocs resp
) => HasDocs (API capture query req resp) where
  docs (OneOf (x::[])) = docs x
  docs (OneOf (x::y::xs)) = docs x ++ "\n" ++ oneOf (y::xs)
\end{minted}

Now \mintinline{idris}{docs userAPI}, assuming that the universes implement
the \mintinline{idris}{HasDocs} interface, will evaluate to a very basic form of documentation:
\begin{verbatim}
# /users/<user_id:CINT>
## GET
 Response type: USER
# /users/find?email=<QSTRING>
## GET
 Response type: LIST USER

@@

\subsection{An interpreter for an HTTP Server}\label{sec:server}
Now that we've shown that it is easy to manipulate APIs because they're just datatypes, it's time for a more complicated example. Though it is cool that we can generate documentation from an API description, the useful part that we want to implement is an HTTP Server.

\subsubsection{A type for web applications}
Lets first ask the question how a web server should look like in Idris. In the Servant paper, they describe the type of http handlers in the \texttt{wai} library, which looks like \cite{servant}:
\begin{minted}{haskell}
type Application = Request -> IO Response
\end{minted}.

We will more-or-less use this representation to model our HTTP handlers. they are  simply a function from request to response.  The actual implementation of a web server based on this model is out of scope for now, but it would be interesting to port something like Haskells' \texttt{warp} (which impelements this concept) to Idris.

Instead of having two abstract types \mintinline{idris}{Request} and \mintinline{idris}{Response}, we will assume for simplicity that a \mintinline{idris}{Response} is just a \mintinline{idris}{String}, and that a \mintinline{idris}{Request} is an \mintinline{idris}{URL}, and a requestBody of type \mintinline{idris}{Maybe String}.

\begin{minted}{idris}
Application : Type
Application = (url : URL) -> (requestBody : Maybe String) -> IO String
\end{minted}

Just like in the Servant paper, we will need to modify this type a little bit more to suit our needs. We need to be able to express that an \mintinline{idris}{Application} does not match in case of a routing mismatch, such that we can try another route encompassed in \mintinline{idris}{OneOf}.

Instead of returning  a value of \mintinline{idris}{IO String}, we can return a value of \mintinline{idris}{Maybe (IO String)}, where \mintinline{idris}{Just} signals a route match, whilst \mintinline{idris}{Nothing} signals a route mistmatch such that we can try another route.

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{minted}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\makeatletter
\renewcommand*{\thelisting}{\thesection.\arabic{listing}}
\@addtoreset{listing}{section}
\makeatother
\author{Arian van Putten \\ a.vanputten@uu.nl}
\title{Servis: A dependently typed DSL for web APIs}
\begin{document}
\maketitle
\begin{abstract}
In this paper we describe a DSL for web APIs named Servis. Servis is  inspired by Servant, a haskell type-level DSL for web APIs. We identify several issues with the current implementation of Servantm and show how Servis solves these by using features of the dependently typed language called Idris.

Specifically, we show how dependent types can help use describe dependencies within an API, how we can check at  compile time that routes in an API dont overlap, and how dependent pairs can let us verify that status codes described in an API are adhered to.

@@

\section{Introduction}
Servant is an extensible type-level DSL for Haskell, that tries to make web APIs \emph{first-class}~\cite{servant}. A web API should be a thing that can be named, passed to, or returned by functions. Just like any other term in a language.

It does this by describing the DSL at the type-level, and then uses \emph{type families} (type-level functions) to be able to perform computations on the API specifications.  By using these type families, we are able to write different \emph{interpretations} of an API\@.  Example of interpretations are: generating documentation from a specification, generating types of a server implementation, or even generating a client that can interfact with the API\@.

In this paper, We will describe Servis, a DSL  similar to  Servant written in Idris. We will point out various issues with the current Servant DSL,
and then show how we can solve this in Idris.

\subsection{An introduction to Servant}
\label{sec:servant_intro}
We will now give a very short overview of the internals of Servant. We will start by describing the type-level DSL of servant.

Servant lives in the type-level, and defines its language as part of the kind of types named \mintinline{idris}{*}, which is an open kind. This has been done on purpose so that the language becomes extensible. We can always add new constructs to the DSL by creating new data definitions.

@@

data Get (ctypes :: [*]) (t :: *)
data Post (ctypes :: [*]) (t :: *)
data JSON
data PlainText

type Echo = "echo"
:> ReqBody ’[PlainText] String
:> Get
’[PlainText] String
\end{minted}
\caption{The Servant DSL and an example of an API definition}
\label{lab:servant}
\end{listing}

It then uses type classes and type families to describe operations on this type-level DSL. For example,
a documentation generation type class exists:
\begin{minted}{haskell}
docs :: HasDocs api => Proxy api -> String
\end{minted}

A more exotic example is the server type class, that transforms an API into a web server:
\begin{minted}{haskell}
class HasServer api where
  type Server api :: *
  route :: Proxy api -> Server api -> RoutingApplication

serve :: HasServer api => Proxy api -> Server api -> Application

echo :: Server Echo 
echo txt = return txt

main :: IO ()
main = run 8080 (serve (Proxy Echo) echo)
\end{minted}

\subsection{Why idris}
Idris is a language that in syntax is rather similar to Haskell, but has a more powerful type system. Idris is a dependently-typed language. This means that types are first-class citizens. We can pass types to functions, let functions create new types, and let types depend on other value-level terms.

Though haskell has type-level functions (called type-families), it does not allow these functions to depend on term-level values. There is a distinct boundary between type-level and term-level, whilst the two are merged in Idris. This makes doing computations with types a lot more natural. No special constructs are needed like special syntax for type-families; you can just write functions like you would normally do with term-level values. It also allows us to do more powerful things like letting types actually depend on runtime user input.

@@

TODO reword
\begin{itemize}
  \item We point out several problems and limitations of the original Servant DSL
  \item We show a port of servant to Idris, a dependently-typed language
  \item We show how new techniques from Idris can help mitigate some of the problems that Servant has.
\end{itemize}

\section{Limitations of Servant}\label{sec:issues}
\subsection{Term and type-level are not unified in Haskell}
As explained in the introduction, the Servant DSL lives in the type-level. In Haskell, types and terms are separated. One can lift terms into the type-level using GHC extensions like \texttt{XDataKinds}.  On the term level, Haskell has many great libraries and data structures that can be used. For example, all kinds of string manipulations. It would be useful to be able to use these functions to manipulate APIs, making them truly first-class. But there is no way to lift existing term-level functionality to the type-level, which makes an API description only first-class on the type level.

For example, a useful function would be \\ \mintinline{haskell}{path "/users/friends/etc" = "users" :> "friends" :> "etc"} to reduce noise in the DSL\@. But this is currently impossible to write because Strings in the type level are of kind \mintinline{haskell}{Symbol} and not \mintinline{haskell}{String} so useful string manipulation functions like \mintinline{haskell}{splitOn "/"} will not work. This would mean that we would have to re-implement all kinds of utility functions on the type-level.

\subsection{A kind of APIs}
As mentioned in Section~\ref{sec:servant_intro}, a deliberate choice has been made to let the Servant DSL live in the open kind \mintinline{haskell}{*}. Though this makes the DSL extensible, it also makes the DSL fragile and prone to error.

@@

The problem is hard to spot, but the user wrote \mintinline{haskell}{->} instead of \mintinline{haskell}{:>}.  This code compiles fine though. Because the Servant DSL lives in kind \mintinline{haskell}{*}, \mintinline{haskell}{:>} is of kind \mintinline{haskell}{* -> * -> *}. The type constructor of functions, \mintinline{haskell}{->} is also of kind \mintinline{haskell}{* -> * -> *}, so there is no way for the compiler to tell that we used the wrong operator here.

@@

Also, as seen in Listing~\ref{lab:servant}, the Servant DSL is split up in several semantic portions.  For example, \mintinline{haskell}{:>} should take an \mintinline{haskell}{item} on the left-hand side, which describes either a constraint like \mintinline{haskell}{ReqBody} or a path-piece, whilst the right-hand side should be an \mintinline{haskell}{api}.
These semantic portions, however,  are implict and unchecked, so we can create totally non-sensical API types like \mintinline{haskell}{type Nonsense = (Int :>Int) :<|> ("hey ":> Int)} that have no interpretation at all.

@@

Another implict rule is that a chain of \mintinline{haskell}{:>} should always end in a \mintinline{haskell}{Get} or a \mintinline{haskell}{Post}. So this should be invalid: \mintinline{haskell}{type API = "a" :> ReqBody [JSON] User}.  This can also not be checked if we do not add more fine-grained kinds to the DSL.

\subsection{API Interdependencies}
\label{sec:inter}
Sometimes, one would like to describe even more type-safety than is currently possible in Servant. Maybe we want to let the output-type of a request depend on a query parameter.
A usecase could be the following: We have a route \texttt{"/users/list?limit=100}", and we want to make sure that a handler never returns more than 100 users.  Thus the return type of the handler should be a \mintinline{idris}{BoundedList 100 User}. Or more generally, the return type should be \mintinline{idris}{BoundedList limit User}, where \mintinline{idris}{limit : Nat} comes from parsing the URL.  The full type of the API will thus depend on some runtime value. This is a textbook example of dependent types, and is currently not possible to implement in Haskell.
However, with Idris this does become possible.

\subsection{Overlapping routes}
Another problem with Servant is that there is no protection for overlapping routes.
If two routes resolve to the same path, Servant makes a left-biased choice~\cite{servant}.
It would be useful that instead of making a design choice where some routes are shadowed without warning, the compiler could inform the user about a potential mistake.

\begin{listing}
\begin{minted}{haskell}
  type API  = "users" :> Capture "args" String :> Get [JSON] User
         :<|> "users" :> "favorites" :> GET [JSON] User
\end{minted}
\caption{An example of overlapping routes. The capture will be chosen if args equals "favorites"}
\end{listing}

\subsection{Explicit status codes}
It is great that Servant lets us give a presice description about the structure and types of our web APIs. However, it does now allow us to describe any form of errors within the API types.
It is important for a developer to know how an API call could fail, such that he can handle it sufficently. Also, when generating documentation from an API type, it one would like to have this information at hand to inform the users of your API on what kind of errors they can expect.

This would not be hard to add to Servant. A type-level list of error types paired with the return type of an API endpoint would suffice. However, it would also be interesting to explore if we can enforce that the code that implements a server for an API, actually only returns the error codes that are documented.

@@

 In Idris, we can define this as the following interface:
 \begin{listing}
 \begin{minted}{idris}
interface Universe (u : Type) where
  el : u -> Type
 \end{minted}
   \caption{Universe interface in Idris}
 \end{listing}

@@

--  el api evaluates to  Int -> IO User
handler : el api
handler = getUser
 \end{minted}
\subsection{The DSL}
So what does this type \mintinline{idris}{API} look like? An API DSL should allow us to define routes, to which we can attach handler types.
\subsubsection{Handler}\label{sec:handler}
Before we will look at routes, we must decide what an endpoint should look like.
We should be able to handle the HTTP verbs. We will only implement \mintinline{idris}{GET} and \mintinline{idris}{POST}, because the others are very similar.
For a \mintinline{idris}{GET}, we should be able to specify a return type of the handler,  and for a \mintinline{idris}{POST}, we should be able to
specify a request body type and a return type.

@@

However, once we start implementing an actual server in Section~\ref{sec:server}, we will discover that this definition is not ideal. What if we would want to write a function that needs to decide its behaviour based on the \mintinline{idris}{responseType} of a \mintinline{idris}{GET}?
For example, we have a piece of functionality in our code base, where we need to decide how to encode a response body to JSON, based on the value of \mintinline{idris}{responseType}. Lets try to implement such a function \mintinline{idris}{selectEncoder} and see what road blocks we hit:

@@

userEncoder : Encoder
userEncoder = selectEncoder User
\end{minted}

Sadly enough, this code will give us a compiler error. The problem is that in Idirs, we cannot pattern match on values of \mintinline{idris}{Type}. 
However, this exact problem can be solved with universes We can define a data type \mintinline{idris}{ResponseUniverse : Type} which names
the \mintinline{idris}{User} type as a data constructor \mintinline{idris}{USER : ResponseUniverse}. We can then pattern-match on \mintinline{idris}{ResponseUniverse} instead of \mintinline{idris}{Type}, and whenever we need the \mintinline{idris}{User} type, we can just call \mintinline{idris}{el USER : Type}.

@@

We now change the definition of \mintinline{idris}{Handler} to be parameterised over a request universe (\mintinline{idris}{req : Type}) and a response universe (\mintinline{idris}{res : Type}):
\begin{listing}
  \begin{minted}{idris}
data Handler : (req : Type)  -> (res : Type) -> Type where
  GET : (responseType : res) -> Handler req res
  POST : (requestType : req) -> (responseType : res) -> Handler req res

@@

\subsubsection{Path and PathPart}
Of course we do not only get request information from the request body,
we also get information about a request by parsing it's request URL. This
is where \mintinline{idris}{Path}s and \mintinline{idris}{PathPart}s come into play. 
When we talk about path parts, we mean sections of a path template that convey parsable information.
For example, the following path template has three path parts: \texttt{/users/\string{user\_id\string}?search=\string{query\string}}. A constant piece
in the path \texttt{users}, a variable capture \texttt{user\_id} and a query param \texttt{search}. This
decomposition of a path template can be written down as a datatype. In this datatype we do not only
want to describe the names of the parts of these paths, but also their types. Hence, we will parameterize
over a universe similarly to how we defined \mintinline{idris}{Handler}.
\begin{minted}{idris}

data PathPart : (capture : Type) -> (query : Type) -> Type where
  ||| A constant piece of text
  ||| @ path the path part
  Const : (path : String) -> PathPart capture query
  ||| A capture of a variable name of a specific type
  ||| @ name  the name of the variable
  ||| @ type  the type of the variable
  Capture : (name : String) -> (type : capture) -> PathPart capture query
  ||| A query param of a variable name of a specific type
  ||| @ name  the name of the variable
  ||| @ type  the type of the variable
  QueryParam : (name : String) -> (type : query) -> PathPart capture query

@@

The path \texttt{/users/\string{user\_id\string}?search=\string{query\string}}, can be described as follows with these new data types:
\begin{minted}{idris}
   Const "users"
:> Capture "user_id" INT
:> QueryParam "search" STRING
:> Outputs (GET USER)
\end{minted}

The universe implementation is straightforward for paths, except for
that we do not want to have spurious arguments in case when we encounter
a \mintinline{idris}{Const}: 
\begin{minted}{idris}
( Universe capture
, Universe query
) => Universe (PathPart capture query) where
  el (Const path) = ()
  el (Capture name type) = el type
  el (QueryParam name type) = el type

@@

\subsubsection{The API Datatype}
Finally, we define an \mintinline{idris}{API} as an non-empty list of \mintinline{idris}{Path}s that can be chosen from.

@@

The universe implementation of an \mintinline{idris}{API} is the type of a list of handlers.
Every handler is of a different type though, so instead of using an ordinary list,
we use an heterogenous list, where each element \mintinline{idris}{x} in \mintinline{idris}{xs}
is of type \mintinline{idris}{el x}.
\begin{minted}{idris}
( Universe capture
, Universe query
, Universe req
, Universe res
) => Universe (API capture query req res) where
  el (OneOf xs) = HVect (map el xs)
\end{minted}

\subsection{An example API}
Using this DSL, we will now describe an API for interacting with users in a database.
We will start off with defining our domain objects, and interactions we can perform on them.
These definitions can be found in Listing~\ref{code:domain}.

\begin{listing}[H]
\begin{minted}{idris}
record User where
  constructor MkUser
  id : Int
  name : String
  email : String
  password : String

findUserByEmail : String -> IO (List User)
getUsers : (limit : Nat) -> IO (List User)
addUser : User -> IO User
\end{minted}
\label{code:domain}
\caption{The User object and its interactions}
\end{listing}

Next, we will describe the HTTP API for interacting with these users.  This includes
the need of coming up with universes for query params, request body, and response. We will use this example in the upcoming sections
to show the behaviour of different interpreters.

% TODO MAKE A GOOD PLACEMENT FOR THIS AND MAYBE SHORTER.
% TODO DESCRIBE EACH STEP. BREAK IT UP!!
\begin{listing}[H]
\begin{minted}{idris}
data QueryUniverse
  = QSTRING
  | QINT
  | QNAT
Universe QueryUniverse where
  el QSTRING = String
  el QNAT = Nat

@@

userAPI : API QueryUniverse CaptureUniverse ReqResUniverse ReqResUniverse
userAPI = OneOf
  [ Const "users" :>
    Const "find" :>
    QueryParam "email" QSTRING :>
    Outputs (GET (LIST USER))
  , Const "users" :> Outputs (POST USER)
  , Const "users" :>
    QueryParam "limit" QNAT :>
    Outputs (GET (LIST USER))
  ]

-- set the handlers
handlers : el userAPI
handlers =
  [ findUserByEmail
  , addUser
  , getUsers
  ]

@@

\subsection{A simple interpreter: Documentation generation}
Writing an interpreter for generating documentation is straightforward. Our API description isn't some type-level definition, it's just a datatype that can be passed to functions and can be manipulated.

We start by introducing an interface for docs, which just tells us how to turn something into a string:
\begin{minted}{idris}
interface HasDocs a where
  docs :: a -> String
\end{minted}
Now, we just need to define an implementation for \mintinline{idris}{API} for this interface and we are done. To implement this, we will of course also needs to implement the interface for \mintinline{idris}{Handler}, \mintinline{idris}{PathPart} and \mintinline{idris}{Path}.

To get an idea on how this would work, we will show the implementation of \mintinline{idris}{Handler} and \mintinline{idris}{API}:
\begin{minted}{idris}
(HasDocs req, HasDocs res) => HasDocs (Handler req res) where
  docs (GET responseType) =
    "## GET\nResponse type: " ++ docs responseType

  docs (POST requestType responseType) =
    "## POST\nRequest type: "  ++ docs requestType ++ "\n" ++
    "Response type: " ++ docs responseType ++ "\n"
\end{minted}

% TODO add text
\begin{minted}{idris}
( HasDocs capture
, HasDocs query
, HasDocs req
, HasDocs resp
) => HasDocs (API capture query req resp) where
  docs (OneOf (x::[])) = docs x
  docs (OneOf (x::y::xs)) = docs x ++ "\n" ++ oneOf (y::xs)
\end{minted}

Now \mintinline{idris}{docs userAPI}, assuming that the universes implement
the \mintinline{idris}{HasDocs} interface, will evaluate to a very basic form of documentation:
\begin{verbatim}
# /users/find?email=<QSTRING>
## GET
 Response type: LIST USER

@@

\subsection{An interpreter for an HTTP Server}\label{sec:server}

Now that we have shown a small example on how to do computations on API descriptions,
it is time for a less contrived example. We would like to implement handlers for our API and then serve these on the web.

\subsubsection{A type for web applications}
Before we write such an interpreter, lets first ask the question what a web server should look like in Idris. In the Servant paper, the authors describe the type of http handlers in the \texttt{wai} library, which looks like \cite{servant}:
\begin{minted}{haskell}
type Application = Request -> IO Response
\end{minted}.
This is a natural way to think of an HTTP application. A request comes in, and we apply a function of type \mintinline{idris}{Application} to get a response, which can then be sent back.

We will more-or-less use this representation to model our HTTP handlers.An actual implementation of a web server based on this model is out of scope for now, but it would be interesting to port something like Haskells' \texttt{warp} (which impelements this concept) to Idris.

Instead of having two abstract types \mintinline{idris}{Request} and \mintinline{idris}{Response}, we will assume for simplicity that a \mintinline{idris}{Response} is just a \mintinline{idris}{String}, and that a \mintinline{idris}{Request} is an \mintinline{idris}{URL}, and a requestBody of type \mintinline{idris}{Maybe String}. This leads us to the following definition for \mintinline{idris}{Application}:

\begin{minted}{idris}
Application : Type
Application = (url : URL) -> (requestBody : Maybe String) -> IO String
\end{minted}
% TODO reword a bit
